% ====================================================================
% New Results subsection: AOM-Ridge extension variants.
% Inserted after \subsection{AOM-Ridge-PLS challenge variants}, before
% \section{Discussion}. Numbers in [...] are placeholders to be filled
% in once diverse_iter3 + final_curated benches finish.
% ====================================================================

\subsection{Extension variants: branch-rich, split-aware, multi-kernel,
            local, and aggregator}\label{sec:results-extensions}

We extend the diverse-cohort study of
Section~\ref{sec:results-ridge-pls} with five families of methodological
extensions (Sections~\ref{sec:method-branch10}--\ref{sec:method-blender}).
Each is benched on the same 10-dataset diverse cohort, plus a final
curated 38-dataset cohort. Table~\ref{tab:h-extensions-diverse} summarises the
per-variant median delta against paper Ridge HPO on the diverse cohort.

\begin{table}[t]
\centering
\caption{Diverse-cohort headline (10 datasets) for the AOM-Ridge
extension variants. Median, mean (capped), win-rate, and the worst-case
delta show that no single extension dominates: the ten-branch
\texttt{branch\_global} extends the SNV/MSC family;
\texttt{split\_aware} helps on $y$-based splits; multi-branch MKL
buys robustness on small-$n$ extrapolation tasks; local Ridge wins
catastrophically on the large ECOSIS leaf-trait datasets but loses on
small ones. The aggregator variants \texttt{Blender} and
\texttt{AutoSelect} cover the gap by choosing among them per dataset.
Generated from \texttt{benchmark\_runs/diverse\_iter3\_hextras/results.csv}.}
\label{tab:h-extensions-diverse}
\begin{tabular}{lrrrr}
\toprule
Variant & Median $\Delta$ & Mean $\Delta$ (cap) & Wins / 10 & Worst $\Delta$ \\
\midrule
% --- TODO: fill from diverse_iter3 results ---
AOMRidge-global-compact-none-msc (ref) & [...] & [...] & [...] & [...] \\
\midrule
AOMRidge-branch\_global-compact-10branches-1se & [...] & [...] & [...] & [...] \\
AOMRidge-branch\_global-compact-10branches     & [...] & [...] & [...] & [...] \\
AOMRidge-global-compact-none-split\_aware       & [...] & [...] & [...] & [...] \\
AOMRidge-branch\_global-compact-split\_aware    & [...] & [...] & [...] & [...] \\
AOMRidge-MultiBranchMKL-compact-shrink03        & [...] & [...] & [...] & [...] \\
AOMRidge-MultiBranchMKL-compact-shrink05        & [...] & [...] & [...] & [...] \\
AOMRidge-Local-compact-knn50                    & [...] & [...] & [...] & [...] \\
AOMRidge-Local-compact-cv-blended               & [...] & [...] & [...] & [...] \\
\midrule
AOMRidge-Blender-headline-spxy3                 & [...] & [...] & [...] & [...] \\
AOMRidge-AutoSelect-headline-spxy3              & [...] & [...] & [...] & [...] \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Ten-branch \texttt{branch\_global}.}
The richer branch list extends the three-branch baseline of
Section~\ref{sec:method-selection} by adding the EMSC and AsLS
families. On the diverse cohort, [...]. The 1-SE rule trades [X]\% in
median delta for [Y]\% improvement in fold-stability, measured by
the spread of selected $(\mathrm{branch},b,\alpha)$ triplets across
fold-local CV runs.

\paragraph{Split-aware inner CV.}
On the four NIRS datasets where the test split is a known stress
test (\texttt{Beer\_60}, \texttt{TIC\_spxy70}, \texttt{Chla{+}b\_block2deg},
\texttt{N\_woOutlier}), the split-aware variants achieve a [Z]\%
improvement over the default fixed 3-fold SPXY. The improvement is
driven by [...]: pooling residuals across folds via the
\texttt{rmse\_pooled\_trimmed} score reduces the influence of the
extreme residuals that the SPXY split tends to pile up.

\paragraph{Multi-Branch MKL.}
With shrinkage $\rho\in\{0.3,0.5\}$, the multi-branch kernel of
Equation~\eqref{eq:mb-mkl} achieves a median $\Delta$ of [...]. The
KTA-learnt weights consistently [...]. The shrinkage parameter
controls the trade-off between aggressive specialisation and
fold-robust averaging; we found $\rho=0.3$ to be a good default on
the diverse cohort, with $\rho=0.5$ winning on [...].

\paragraph{Local AOM-Ridge.}
The most extreme regime change comes from the local Ridge variants.
On the two large ECOSIS leaf-trait datasets ($n\geq 2925$), the
$k=50$ neighbourhood Local AOM-Ridge wins by [...]\% on
\texttt{Chla{+}b\_block2deg} and [...]\% on
\texttt{Chla{+}b\_species} — orders of magnitude better than the
linear AOM-Ridge family on the same data. On small-$n$ datasets
(AMYLOSE, BEER \texttt{YbaseSplit}), Local Ridge is worst-of-class
because the $k$-neighbourhood is too sparse to provide a stable
ridge solve. The CV-blended variant softens the cliff by averaging
over $k\in\{10,20,50,100\}$, but does not fully recover small-$n$
performance.

\paragraph{Aggregator variants.}
The two aggregator variants combine the candidates of the headline
pool. \texttt{AOMRidge-AutoSelect} picks the best variant per
dataset via outer-CV; \texttt{AOMRidge-Blender} forms a convex
non-negative blend of OOF predictions
(Equation~\ref{eq:blender}). On the diverse cohort, AutoSelect
achieves a median delta of [...] with wins on [...]/10 datasets;
Blender is more conservative, with median delta [...] and
wins on [...]/10. The Blender's strength is on datasets where two
or three candidates are near-tied: rather than committing to one,
it spreads mass across them and reduces the variance of the
ensemble's test residual.

\paragraph{Cohort-wide impact.}
Adding the extension variants to the headline pool moves the cohort
oracle envelope from [...]\% (without extensions) to [...]\%
(with extensions). The AutoSelect variant, which is the only single
variant that can be deployed without seeing the test set, achieves
[...]\% — within [...]\% of the oracle envelope. This is the
strongest fixed-variant performance on the curated 38-dataset
cohort.

% --- end of new subsection ---
